---
title: "36-402 Homework 3"
author: "James \"Morgan\" Hawkins"
date: "Friday Feb 10, 2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
```


## Problem 1


### Problem 1 (c)

```{r}
rx <- function(b){
  if(b == 0){
    return(runif(1,0,4))
  }else{
    return(rnorm(1,-2,2))
  }
  
}

set.seed(99)
sample.z <- rbinom(10000,1,.5)
sample.x <-sapply(sample.z,rx)

mean(sample.x)
sd(sample.x)^2
```


We expect our sample to have a mean of 0 and a variance of 6.666. After drawing 10,000 smaples from the distribution of x we get a sample mean of .0100 and a sample variance of 6.6742. This reinfoces our answers for part a and b.


## Problem 4


### Problem 4 (a)


```{r}
1198/(1493+1198)

557/(557+1278)

```


44.52% of male applicants were accepted to Berkeley while only 30.35% of female applicants were accepted. 


### Problem 4 (b)


```{r}
ucb = data.frame(UCBAdmissions)
ucb
```


```{r}
calc_acceptance_rates <- function(df, department){
  df = df[df$Dept == department,]
  df = df[order(df$Admit),]
  df = df[order(df$Gender),]
  return(c(
    "Male" = df[1,4]/(df[1,4] + df[2,4]),
    "Female" = df[3,4]/(df[3,4] + df[4,4])
            ))
}

for(d in unique(ucb$Dept)){
  temp.admiss = (calc_acceptance_rates(ucb,d))
  
  #print(temp.admiss)
  print(d)
  if(temp.admiss[[1]]>temp.admiss[[2]]){
    print("male higher")
  }else{
    print("female higher")
  }
  
}

```


The female acceptance rate is higher in 4 departments (A, B, D, F) and male acceptance rate is higher in 2 departments (C, E)


This can be considered a near example of simpson's paradox because our conclusion changes based on how we look at our data. In part A we looked at overall acceptance, but in part b we looked at the proportion of departments that accept male applicants at a higher rate. 


### Problem 4 (c)


Y -- accepted? bool

X -- m/f [1 = female]

Z -- department -- categorical


```{r}

calc_acceptance_rates <- function(df, department){
  df = df[df$Dept == department,]
  df = df[order(df$Admit),]
  df = df[order(df$Gender),]
  return(c(
    
    df[1,4]/(df[1,4] + df[2,4]),
    df[3,4]/(df[3,4] + df[4,4])
    
    ))
}
  
department.acceptances = data.frame("Gender" = c("Male","Female"))
for(d in unique(ucb$Dept)){
  
  #append(department.acceptances, c(calc_acceptance_rates(ucb,d)) )
  department.acceptances[,paste("Z",d,sep = "=")] = calc_acceptance_rates(ucb,d)
}

department.acceptances


```






covariate - country of origin

treatment - drinks tea

outcome - if you live past 40







